---
title: 'Learning material'
output: 
  html_document:
    css: custom.css
    toc: no
---
```{r, include=FALSE}
library(htmltools)
suppressPackageStartupMessages(library(htmltools))
source('utils.R')
source('reveal.R')
```

\  

This page contains lecture slides that we deliver during our in-person courses and short how-to videos to help support you whilst learning R. This material should be used in conjunction with the [Introduction to R book](https://alexd106.github.io/Rbook). 

\  

## Lecture slides

Follow the links below for PDF versions of lecture slides we deliver during our in-person courses.

\  

\  

```{r, reveal, echo=FALSE, eval = isFALSE(show_lecture)}
cat("Lectures will be released as the course progresses")
```


```{r, echo=FALSE, eval = isTRUE(show_lecture)}
thumbnail("Introduction to R", "images/r_lecture_thumb.png",
          "lectures/introduction_to_r_2020.pdf")
```

```{r, echo=FALSE, eval = isTRUE(show_lecture)}
thumbnail("Data in R", "images/data_in_r_thumb.png",
          "lectures/getting_the_data_in_r.pdf")
```

```{r, echo=FALSE, eval = isTRUE(show_lecture)}
thumbnail("Simple Statistics in R", "images/stats_r_thumb.png",
          "lectures/statistics_in_r.pdf")
```

\  

\  

## How-to

Short how-to videos to get you started with some important topics. 

\  

<iframe src="https://player.vimeo.com/video/415501284" width="640" height="436" frameborder="0" allow="autoplay; fullscreen" allowfullscreen></iframe>

\  

<iframe src="https://player.vimeo.com/video/416391353" width="640" height="381" frameborder="0" allow="autoplay; fullscreen" allowfullscreen></iframe>

\  


<iframe src="https://player.vimeo.com/video/416877495" width="640" height="381" frameborder="0" allow="autoplay; fullscreen" allowfullscreen></iframe>

\  

<iframe src="https://player.vimeo.com/video/416902337" width="640" height="381" frameborder="0" allow="autoplay; fullscreen" allowfullscreen></iframe>

\  

<iframe src="https://player.vimeo.com/video/417242115" width="640" height="480" frameborder="0" allow="autoplay; fullscreen" allowfullscreen></iframe>

\  

<iframe src="https://player.vimeo.com/video/416743698" width="640" height="381" frameborder="0" allow="autoplay; fullscreen" allowfullscreen></iframe>

\  

